---
title: "DATA-413/613 HW 3: Lists and Iteration"
author: "Evan Bowman"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: no
    toc_depth: 4
urlcolor: "blue"
params:
  solutions: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = params$solutions,
  eval = params$solutions,
  fig.align = "center",
  fig.height = 3,
  fig.width = 5
)
```

# Instructions {.unnumbered}

1.  Clone this homework repo to your homework directory as a new repo.
2.  Rename the starter file under the analysis directory as `hw_03_yourname.Rmd` and use it for your solutions.\
3.  Modify the "author" field in the YAML header.\
4.  Stage and Commit R Markdown and HTML files (no PDF files).\
5.  **Push both .Rmd and HTML files to GitHub**.\

-   Make sure you have knitted to HTML prior to staging, committing, and pushing your final submission.\

6.  **Commit each time you answer a part of question, e.g. 1.1**\
7.  **Push to GitHub after each major question**\
8.  When complete, submit a response in Canvas

-   Only include necessary code to answer the questions.

-   Most of the functions you use should be from the tidyverse. Too much base R will result in point deductions.

-   Use Pull requests and or email to ask me any questions. If you email, please ensure your most recent code is pushed to GitHub.

-   **Learning Outcomes**:

    -   Manipulate vectors in base-R syntax.
    -   Apply iterations with for loops.

# Grading Rubric {.unnumbered}

| Question.Part: | Points | Topic                                          |
|----------------|--------|------------------------------------------------|
| 1.1            | 1.00   | Load the Data                                  |
| 1.2            | 1.00   | Correct length and Names                       |
| 1.3            | 1.00   | Values are Removed from the list               |
| 1.4            | 2.00   | Correct Posterior Mean added to scores         |
| 1.5            | 2.00   | Correct Column Means added to list             |
| 1.6            | 2.00   | Extracted values are correct                   |
| 2.1a           | 2.00   | Proper Function structure                      |
| 2.1a           | 3.00   | Complete Error checks                          |
| 2.1b           | 1.00   | Correct Output                                 |
| 3.1            | 1.00   | Single function with Element names and types   |
| 3.2            | 3.00   | For Loop is initialized and operates correctly |
| 3.3            | 3.00   | Correct map output and Plot                    |
| 3.4            | 1.50   | Correct Output for pvec_f                      |
| 3.5            | 1.50   | Correct Output for pvec_m                      |
| 3.6            | 1.00   | Extra Credit                                   |
| **Total**      | **25** | plus 1 pt Extra Credit                         |

```{r}
library(tidyverse)
```

# Working with Lists

Because of their generality, lists (or list-like objects) are often the output of many statistical procedures in R. The file `fpout.RDS` in the data folder contains sample output from using [fitPoly](https://cran.r-project.org/package=fitPoly), a statistical method to quantify properties of locations on the [genome](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5915555/).

1.  Use `read_rds()` and a relative path to read this data into R.

```{r}
fitPoly <- read_rds("../data/fpout.RDS")
```

2.  Use appropriate functions to obtain the length of the list and then the vector of the names of the elements in the list.

```{r}
# Length of List
length(fitPoly)

# Vector of names of the elements
names(fitPoly)
```

3.  The `diploscores` element does not provide any information. Remove it from the list in one step.

```{r}
fitPoly <- fitPoly[-5]

names(fitPoly)
```

4.  The `scores` element contains the output most users would want. The variables in `scores` called `P0`, `P1`, `P2`, `P3`, `P4`, `P5`, and `P6` contain "posterior probabilities" for each individual for values `0`, `1`, `2`, `3`, `4`, `5`, and `6` (respectively).\

-   A quantity useful in Bayesian analysis is called the "posterior mean" which in this case is calculated as follows:
    -   $\text{posterior_mean] = (P0 * 0) + (P1 * 1) + (P2 * 2) + (P3 * 3) + (P4 * 4) + (P5 * 5) + (P6 * 6)$.
-   Calculate the posterior mean for each individual and add this as a new variable in the `scores` data frame.

```{r}
fitPoly$scores <- fitPoly$scores %>%
  mutate(posterior_mean = (P0 * 0) + (P1 * 1) + (P2 * 2) + (P3 * 3) + (P4 * 4) + (P5 * 5) + (P6 * 6), .after = P6)
head(fitPoly$scores)
```

5.  Operating on elements within a list.

-   Use a {purrr} function to identify the names of the columns/variables in the `scores` data frame that are *not* of type double.\

```{r}
names(keep(fitPoly$scores, ~ !is_double(.)))
```

-   Add a new atomic vector element called `col_means` **to the list** that contains just the column means of all the *double* variables in the `scores` data frame.

```{r}
fitPoly$score_col_means <- map_dbl(fitPoly$scores[-1:-3], mean, na.rm = T)
fitPoly$score_col_means

```

6.  Extract elements from a list. 

-   Demonstrate three different ways to extract the `col_means` element from the list. The extracted element must *not be a list*.

```{r}
# 1st Method: Extracting with $
fitPoly$score_col_means
typeof(fitPoly$score_col_means)

# 2nd Method: Extract using [[list_number]]
fitPoly[[7]]
typeof(fitPoly[[7]])

#3rd Method: pluck function
pluck(fitPoly, "score_col_means")
typeof(pluck(fitPoly, "score_col_means"))
```

-   Show two ways to extract the third element of `col_means`

```{r}
# 1st Method: Extracting with both $ and []
fitPoly$score_col_means[3]
typeof(fitPoly$score_col_means[3])

# 2nd Method: Pluck function
pluck(fitPoly, "score_col_means", 3)
typeof(pluck(fitPoly, "score_col_means", 3))
```

# For Loops

Consider the recursive sequence defined by $$x_n = x_{n-1} + \frac{|x_{n-3} - x_{n-2}|}{4}.$$ 
In words, element $n$ is the sum of element $n-1$ and the absolute value of the difference between between elements $n-3$ and $n-2$, divided by four. For example, if we let $x_1 = 3$, $x_2 = 1$, and $x_3 = 10$, then $x_4$ is $$ x_4 = 10 + \frac{|3 - 1|}{4} = 10.5.$$

1.  Write a function called `calcn()` that takes as input a vector `x` with the first three elements of this sequence ($x_1$, $x_2$, $x_3$) and an integer `n` denoting the final element of the sequence to calculate.\

-   `calcn(x, n)` should return the value of the `n`^th^ element.
-   Include error checking to ensure both inputs are of the correct length and class and `n` is greater than 0.


```{r}
calcn <- function(x, n) {
  stopifnot(length(x) == 3 & is.numeric(x))
  stopifnot(n > 0 & is.numeric(n))
  
nums <- vector(mode = "integer", length = n)
for (i in seq_along(nums)) {
  if (i <= 3) {
    nums[i] <- x[i]
} else {
  nums[i] <- nums[i-1] + ((abs(nums[i-3] - nums[i-2])) / 4)
  }
}
  return(nums[n])
}
```

For example, in my implementation of `calcn()`, I obtained the following:

```{r}
# From Homework Document
calcn(x = c(2, 4, 3), n = 3L)
calcn(x = c(2, 4, 3), n = 4L)
calcn(x = c(2, 4, 3), n = 5L)
calcn(x = c(2, 4, 3), n = 6L)
calcn(x = c(2, 4, 3), n = 7L)
```

-   **Evaluate your function at the following inputs**:

```{r}
calcn(c(11, 1, 130), 1000L)  
calcn(c(11, 1, 130), 1L)  
calcn(c(7, 3, 20), 8L)  
```

# Question: Lists, For-loops, and `map_*()`

Lists are often used to save simulation output. You can then extract individual elements from the lists using for-loops.

Consider the $t$-test, often used to test whether or not the true mean of a population for some observations is 0. We can use the following code to simulate data from a [Normal (0,1) distribution](https://en.wikipedia.org/wiki/Normal_distribution), and then use a $t$-test to test if the true mean is 0:

```{r}
set.seed(123)
x <- rnorm(n = 10, mean = 0, sd = 1)
tout <- t.test(x)
```

1.  The output of `t.test()` is a list-like object. Use one function to show how many elements are in the list along with their names and class.
```{r}
str(tout)
```

2.  Write a for-loop to perform the following three operations on iteration `i`:  

    1.  Draw 10 random observations from a normal distribution with mean 0 and standard deviation 2.
    2.  Run a $t$-test on these 10 observations.
    3.  Save the output of the $t$-test as the `i`th element in a list called `tlist`.  

-   Set the seed to 1 and run for 1000 iterations.
```{r}
set.seed(1)
tlist <- vector(mode = "list", length = 1000)
for (i in 1:1000) {
 x <- rnorm(10, mean = 0, sd = 2)
 tout <- t.test(x)
 tlist[[i]] <- tout
}
```

3.  Use the appropriate `purrr:map_*` function to extract the sample mean from each test (from 2) and pipe to an appropriate plot to show the sampling distribution of the sample mean.
-   Hint: Make sure the data going into ggplot is a data frame (tibble)
```{r}
tlist %>%
map_dbl(~.$estimate[[1]]) %>%
  tibble(mean=.) %>%
  ggplot(aes(x = mean))+
  geom_histogram(bins = 70) +
  ggthemes::theme_fivethirtyeight()+
  ggtitle("Distribution of Sample Means")
```


4.  The $p$-value is an important quantity in statistics. Use a for-loop to extract the $p$-values from each test in part 2 and combine these $p$-values into a single atomic vector called `pvec_f`. Show the first 6 values.
```{r}
pvec_f <- vector(mode = "double", length = 1000)
for (i in 1:1000) {
  pvec_f[[i]] <- tlist[[i]]$p.value
}
head(pvec_f)
```

5.  Use the appropriate `purrr:map_*` function to extract the $p$-values from each test in part 2 and combine these $p$-values into a single atomic vector called `pvec_m`. Show the first 6 values.
```{r}
pvec_m <- tlist %>%
  map_dbl(~ .$p.value)
head(pvec_m)
```

6.  *Extra Credit* $p$-values have a nice property where, if the null hypothesis is true (i.e., the mean of the observations is actually 0), then the $p$-values follow the uniform distribution.

-   Use the data from `pvec_m` to create a [QQ-plot](https://en.wikipedia.org/wiki/Q–Q_plot) and then interpret the plot with regard to whether the $p$-values exhibit a uniform distribution.\
-   Ensure your plot has appropriate labels for the axes and a title.
-   Include an abline that is dashed and colored red.

